On Devon Allen's Disqualification at the 2022 Track and Field World Championship

### 2023-04-25 Manuscript/manuscript.tex
Went over Prof. Schifano's comments

%2)
Allen's disqualification has been discussed extensively in online communities
such as \url{www.LetsRun.com}, a website that is part message board and part
news. The message board,
which functions similarly to Reddit is very active during important running
events, including during the finals of major events such as
the Olympics and the World Track and Field Championships.  However, it is the
\jy{cite these blog articles formally}
news section of the website that generated articles such as ``Was Devon Allen
Screwed? There's At Least a 99\% Chance That He Was" and ``The Data Keeps Pouring 
In and
It Continues to Look Bad For World Athletics and Great For Devon Allen".  Robert
Johnson, who created and runs the LetsRun website argued that there was an error
with the timing equipment at the 2022 Track and Field World Championship and
cited graphs, simple statistics such as median reaction time, and compared the 
reaction times at the 2022 World Track and Field Championship 
\citep{johnson2022devon}.
\eds{compared the reaction times at the 2022 World Track and Field  Championship 
to what? Also restate here that no formal statistical analyses have been done.}

@@

%4)
\jy{language needs to be formalized and condensed.}
The objective to this paper is to analyze the reaction times at the 2022
World Track and Field Championship to check for any abnormalities, and we approach
this two different ways.  First we try to model the data, which is
the reaction times of every non-disqualification since the 1999 World Championships. 
The model we use is a gamma linear mixed-effects model with a random effect
accounting for the venue (or year) effect. 
The second method is to compare reaction times for athletes who have competed
at additional competitions to the 2022 World Track and Field Championship. 
Our investigation into reaction times is important because reaction times have
been shown to affect the overall time of the race \citep{delalija2008reaction}.
Thus, faster reaction times lead to faster overall times; which is the goal of
every sprinter who steps onto the track.
%\eds{I recommend moving the rest of this paragraph to Section 2.2 as I don't
%think it is critical for readers to know all the details here.}
%First we can compare the reaction times of athletes who
%competed at the World Track and Field Championship in both 2019 and 2022, and 
%second we can compare how athletes performed at the 2022 championship relative 
%to their national/qualifiers meet. In order to reduce any chances
%of athletes improving reaction time, we can also look at data from qualifying meets.
%Almost every athlete had to qualify by proving they are elite at the national level, 
%usually by coming in at least fourth. Thus, data was compiled on athletes who competed
%at the 2022 World Track and Field Championships on what these athletes also ran 
%at their country's national championship. If there was any argument that athletes
%have improved their reaction time from 2019 to 2022, it should now be considered
%nearly null, as the national competitions took place between June and July 2022 
%and the World Championships were in August. The equipment used to measure reaction
%time cannot be guaranteed to be the same from national to international
%competition, but the time span between meets has been reduced considerably.

The rest of this paper is organized as follows. Section~\ref{sec:Data} describes 
how data was collected and begins to detail the generalized linear mixed-effect 
(GLMM) model that is further developed in the Section~\ref{sec:Methods}. The 
signed-rank test that analyzes reaction times for the same athletes is
also described in Section~\ref{sec:Methods}.  Section~\ref{sec:Results}
 provides the results of the two analyses and ends 
by questioning the reaction time barrier.  Lastly, Section~\ref{sec:Discussion}
discusses the impact and limitations of the paper.
\eds{capitalize first letter when referring to a numbered section, table, figure, 
etc. ``Section 4 provides...'' vs. ``The next section provides...''}

@@

There are two types of data that are explored throughout this paper: data from
1999-2022 for every world championship, and data from athletes who competed
the 2022 World Championships and another competition (either 2019 World 
Championship or a 2022 national competition held within two months of the  
2022 World Championships. The first data set is used for the GLMM portion
of the paper, while the second data set is used for the rank-based comparisons.

@@

\subsection{Beyond World Championship}
This section was motivated from a qualitative analysis that started by looking
at how United States athletes reacted at the 2022 United States Track and Field
Championships. From June 23-26 2022, USA held its Track
and Field Championship to decide who to send to the World Championships at Hayward 
Field, the exact same venue as used for the World Track and Field Championship a month 
later.  Thus, a baseline is able to be established for the four USA athletes who 
competed in at least one round of both events: Trey Cunningham, Daniel Roberts, 
Grant Holloway, and Devon Allen. Every athlete reacted faster in all the World 
Track and Field Championship races compared to the USA Track and Field 
Championship races. Devon Allen raced three times at the USA Track and
Field Championship and three times at the World Track and Field  Championship. 
His reaction times were significantly faster in the World Track and
Field Championship compared to the USA Track and Field Championship as were his
teammates.  This comparison suggests that the timing system used between the two
meets may have been meaningfully different to the point where Devon Allen was
disqualified more because of faulty equipment rather than because he reacted
\jy{cross-reference to the table by its label}
too quickly.  Table~\ref{fig:USAvsWorld} highlights the difference in reaction times for
Devon Allen, Trey Cunningham, Grant Holloway, and Daniel Roberts. ``USA" denotes
the USA Track and Field Championships, ``World" denotes the World Track and Field
Championships,   All numbers listed in the table are the reaction
time for the respective athletes in seconds.

\begin{table}
\begin{center}
  \begin{tabular}{c c c c c c c} 
   \toprule
   Athlete & USA H & USA S & USA F & World H & World S & World F \\ [0.5ex] 
   \midrule
   Devon Allen & 0.201 & 0.153 & 0.160 & 0.123 & 0.101 & 0.099 \\ 
   Trey Cunningham & 0.186 & 0.185 & 0.182 & 0.115 & 0.120 & 0.109 \\
   Grant Holloway & 0.192 & 0.190 & N/A & 0.147 & 0.128 & 0.124 \\
   Daniel Roberts & 0.181 & 0.200 & 0.183 & 0.179 & N/A & N/A \\ [0.5ex]
   \bottomrule
  \end{tabular}
  \label{fig:USAvsWorld}
  \caption{Rection times for four United States Athletes at the 2022 USA Track
  and Field World Championships and the 2022 World Track and Field Championship.
  ``N/A" was used for any athlete that did not compete and thus did not register 
  a reaction time.}
  \end{center}
\end{table}

\jy{You have two such datasets in the analysis. Use one paragraph for each.}

\eds{unless target journal has specified otherwise, Table captions should go
above the table. (Figure captions go below the figure)}

@@

\citet{datta2005rank} proposed an extension of the Wilcoxon rank-sum test to
clustered data with subnit-level grouping. The test is designed based on a
within-cluster resampling principle. Consider randomly picking one observation
from each cluster to form a pseudo-sample. Let $X_i^*$ be a random pick from the
$i$th cluster in the pseudo-sample and $\delta_i^*$ its group indicator. The
Wilcoxon rank-sum statistic for the pseudo-sample is
\[
W^* = \frac{1}{n + 1} + \sum_{i=1}^{n} \delta_{i}^{*} R_{i}^{*},
\]
where $R_{i}^{*}$ is the rank of $X_{i}^{*}$ in the psudo-sample.
The test statistic $S$ is the average of $W^*$ averaged over all possible
pseudo-samples conditioning on the observed data and group indicators.
The mean and variance of $S$ under $H_0$ can be derived so that $S$ can be
standardized to form a statistic $Z$ which follows standard normal distribution
asymptotically \citep[p.910]{datta2005rank}. An implementation of this method is
available from the \texttt{clusWilcox.test()} function with
\texttt{method = 'ds'} from R package \texttt{clusrank}
\citep{jiang2017wilcoxon}.
%% In this case, athletes who have more
%% data are generally thought to be faster, as they have made it further in the
%% competition.  Clusters range from two to six, with two indicating that at each
%% competition they only made it to the heats, but a cluster size of six indicates
%% that they made the finals in both competitions.  While the number of times
%% athletes compete is based off their overall time not their reaction time, the
%% two are correlated \citep{delalija2008reaction}.

@@

@@

\jy{Close captions with a period.}
\begin{table}
  \centering
  \begin{tabular}{c c c c c c}
      \toprule
      Dataset & Venue standard error & Int effect est & Int se & Int var & Resid Var \\
      \midrule
      Older Finals & 0.0439 & $-1.901$ & 0.0251 & 0.0019 & 0.0109 \\
      All Finals & 0.0504 & $-1.914$ & 0.0280 & 0.0025 & 0.0112 \\
      Older Pooled & 0.0495 & $-1.853$ & 0.0305 & 0.0024 & 0.0243 \\
      All Pooled & 0.0554 & $-1.869$ & 0.0348 & 0.0031 & 0.0235 \\
      \bottomrule
  \end{tabular}
  \caption{Summary statistics for the GLMM on various data sets}
  \label{tab:Gamma_parameters}
\end{table}

At a size of $n=10,000,000$, the calculated probability of a reaction
time being below 0.1 seconds was as follows for the various different data sets:
\jy{label the table and reference to it; clean it up}
\begin{table}
  \centering
  \begin{tabular}{c c c c} 
   \toprule
   Data Set & P(RxnTime $<$ 0.1) & P(RxnTime $<$ 0.09) & P(RxnTime $<$ 0.08) \\ 
   \midrule
   all finals & $8.56\cdot10^{-4}$ & $4.1\cdot10^{-5}$ & $6.0\cdot10^{-7}$ \\ 
   old finals & $4.39\cdot10^{-4}$ & $1.63\cdot10^{-5}$ & $2.0\cdot10^{-5}$ \\
   all pooled & $6.86\cdot10^{-3}$ & $1.21\cdot10^{-3}$ & $1.3\cdot10^{-4}$ \\
   old pooled & $5.76\cdot10^{-3}$ & $9.93\cdot10^{-4}$ & $1.0\cdot10^{-4}$\\
   \bottomrule
  \end{tabular}
  \caption{Probabilities of observing reaction times less than 0.1, 0.09, and
  0.08 for various data sets}
  \label{tab:Sim_probability}
\end{table}

@@

\jy{motivate first; describe the data (what are the grouping variables in the
  two data sets? use one paragraph for each.}

>>>>

%2)
Allen's disqualification has been discussed extensively in online communities
such as \url{www.LetsRun.com}, a website that is part message board and part
news. The message board,
which functions similarly to Reddit is very active during important running
events, including during the finals of major events such as
the Olympics and the World Track and Field Championships.  However, it is the
news section of the website that generated articles such as \citet{johnson2022was}
and \citet{johnson2022data} from  Robert Johnson, who created and runs 
the LetsRun website.  He argued that there was an error
with the timing equipment at the 2022 Track and Field World Championship and
cited graphs, simple statistics such as median reaction time, and compared the 
reaction times at the 2022 World Track and Field Championship to other competitions
such as the USA Track and Field Championships and other World Championships.
However, their analysis lacked formal analysis. The alleged 99.95\% chance Allen
was wrong came from that the lowest average reaction time for the men's 100 meter,
men's 110 meter hurdle, and women's 100 meters from 2003 to 2022 all occurred at
the 2022 World Championships and Johnson claimed that the probability of this
happening was $\frac{1}{13} \cdot \frac{1}{13} \cdot \frac{1}{13} = 0.05\%$
\citep{johnson2022was}.
\eds{compared the reaction times at the 2022 World Track and Field  Championship 
to what? Also restate here that no formal statistical analyses have been done.}
\of{Cleaned this up a little but still a little messy.}

@@

%4)
\jy{language needs to be formalized and condensed.}
The objective to this paper is to analyze the reaction times at the 2022
World Track and Field Championship to check for any abnormalities, and we approach
this two different ways.  First we try to model the data, which is
the reaction times of every non-disqualification since the 1999 World Championships. 
The model we use is a gamma linear mixed-effects model with a random effect
accounting for the venue (or year) effect. 
Then we compare reaction times for athletes who have competed
at additional competitions to the 2022 World Track and Field Championship.  By
comparing the same athlete across multiple competitions we are able to see if the
2022 data differed significantly from other competitions.
Our investigation into reaction times is important because reaction times have
been shown to affect the overall time of the race \citep{delalija2008reaction}.
Thus, faster reaction times lead to faster overall times; which is the goal of
every sprinter who steps onto the track.
%\eds{I recommend moving the rest of this paragraph to Section 2.2 as I don't
%think it is critical for readers to know all the details here.}
%First we can compare the reaction times of athletes who
%competed at the World Track and Field Championship in both 2019 and 2022, and 
%second we can compare how athletes performed at the 2022 championship relative 
%to their national/qualifiers meet. In order to reduce any chances
%of athletes improving reaction time, we can also look at data from qualifying meets.
%Almost every athlete had to qualify by proving they are elite at the national level, 
%usually by coming in at least fourth. Thus, data was compiled on athletes who competed
%at the 2022 World Track and Field Championships on what these athletes also ran 
%at their country's national championship. If there was any argument that athletes
%have improved their reaction time from 2019 to 2022, it should now be considered
%nearly null, as the national competitions took place between June and July 2022 
%and the World Championships were in August. The equipment used to measure reaction
%time cannot be guaranteed to be the same from national to international
%competition, but the time span between meets has been reduced considerably.

The rest of this paper is organized as follows. Section~\ref{sec:Data} describes 
how data was collected and begins to detail the generalized linear mixed-effect 
(GLMM) model that is further developed in Section~\ref{sec:Methods}. The 
signed-rank test that analyzes reaction times for the same athletes is
also described in Section~\ref{sec:Methods}.  Section~\ref{sec:Results}
 provides the results of the two analyses and ends 
by questioning the reaction time barrier.  Lastly, Section~\ref{sec:Discussion}
discusses the impact and limitations of the paper.
\eds{capitalize first letter when referring to a numbered section, table, figure, 
etc. ``Section 4 provides...'' vs. ``The next section provides...''}

@@

There are two types of data that are explored throughout this paper: data from
1999-2022 for every world championship, and data from athletes who competed
the 2022 World Championships and another competition (either 2019 World 
Championship or a 2022 national competition held within two months of the  
2022 World Championships). The first data set is used for the GLMM portion
of the paper, while the second data set is used for the rank-based comparisons.

@@

\subsection{Beyond World Championship}
This section was motivated from a qualitative analysis that started by looking
at how United States athletes reacted at the 2022 United States Track and Field
Championships. From June 23-26 2022, USA held its Track
and Field Championship to decide who to send to the World Championships at Hayward 
Field, the exact same venue as used for the World Track and Field Championship a month 
later.  Thus, a baseline is able to be established for the four USA athletes who 
competed in at least one round of both events: Trey Cunningham, Daniel Roberts, 
Grant Holloway, and Devon Allen. Every athlete reacted faster in all the World 
Track and Field Championship races compared to the USA Track and Field 
Championship races. Devon Allen raced three times at the USA Track and
Field Championship and three times at the World Track and Field  Championship. 
His reaction times were significantly faster in the World Track and
Field Championship compared to the USA Track and Field Championship as were his
teammates.  This comparison suggests that the timing system used between the two
meets may have been meaningfully different to the point where Devon Allen was
disqualified more because of faulty equipment rather than because he reacted
too quickly.  Table~\ref{fig:USAvsWorld} highlights the difference in reaction times for
Devon Allen, Trey Cunningham, Grant Holloway, and Daniel Roberts. ``USA" denotes
the USA Track and Field Championships, ``World" denotes the World Track and Field
Championships,   All numbers listed in the table are the reaction
time for the respective athletes in seconds.

\begin{table}
\begin{center}
  \caption{Rection times for four United States Athletes at the 2022 USA Track
  and Field World Championships and the 2022 World Track and Field Championship.
  ``N/A" was used for any athlete that did not compete and thus did not register 
  a reaction time.}
  \begin{tabular}{c c c c c c c} 
   \toprule
   Athlete & USA H & USA S & USA F & World H & World S & World F \\ [0.5ex] 
   \midrule
   Devon Allen & 0.201 & 0.153 & 0.160 & 0.123 & 0.101 & 0.099 \\ 
   Trey Cunningham & 0.186 & 0.185 & 0.182 & 0.115 & 0.120 & 0.109 \\
   Grant Holloway & 0.192 & 0.190 & N/A & 0.147 & 0.128 & 0.124 \\
   Daniel Roberts & 0.181 & 0.200 & 0.183 & 0.179 & N/A & N/A \\ [0.5ex]
   \bottomrule
  \end{tabular}
  \label{fig:USAvsWorld}

  \end{center}
\end{table}

\jy{You have two such datasets in the analysis. Use one paragraph for each.}
\of{Split it into 2 sections, do you stil want me to do something else?}

\eds{unless target journal has specified otherwise, Table captions should go
above the table. (Figure captions go below the figure)}
\of{Moved table caption to be above}

@@

\citet{datta2005rank} proposed an extension of the Wilcoxon rank-sum test to
clustered data with subunit-level grouping. The test is designed based on a
within-cluster resampling principle. Consider randomly picking one observation
from each cluster to form a pseudo-sample. Let $X_i^*$ be a random pick from the
$i$th cluster in the pseudo-sample and $\delta_i^*$ its group indicator. The
Wilcoxon rank-sum statistic for the pseudo-sample is
\[
W^* = \frac{1}{n + 1} + \sum_{i=1}^{n} \delta_{i}^{*} R_{i}^{*},
\]
where $R_{i}^{*}$ is the rank of $X_{i}^{*}$ in the psudo-sample.
The test statistic $S$ is the average of $W^*$ averaged over all possible
pseudo-samples conditioning on the observed data and group indicators.
The mean and variance of $S$ under $H_0$ can be derived so that $S$ can be
standardized to form a statistic $Z$ which follows standard normal distribution
asymptotically \citep[p.910]{datta2005rank}. An implementation of this method is
available from the \texttt{clusWilcox.test()} function with
\texttt{method = ``ds'} from R package \texttt{clusrank}
\citep{jiang2017wilcoxon}.
%% In this case, athletes who have more
%% data are generally thought to be faster, as they have made it further in the
%% competition.  Clusters range from two to six, with two indicating that at each
%% competition they only made it to the heats, but a cluster size of six indicates
%% that they made the finals in both competitions.  While the number of times
%% athletes compete is based off their overall time not their reaction time, the
%% two are correlated \citep{delalija2008reaction}.
%% Group is either designated to be 1 or 2, with 1 corresponding to either the
%% national level data or the previous world championship data and 2 corresponding
%% to the 2022 data.

@@

\of{Add something like this to the start of the paragraph?}
We may want to consider using all four data sets for this problem because
comparing all the data to the data excluding 2022 can give us a qualitative
assessment of how impactful 2022 was.

@@

\jy{Close captions with a period.}
\begin{table}
  \centering
  \caption{Summary statistics for the GLMM on various data sets.}
  \begin{tabular}{c c c c c c}
      \toprule
      Dataset & Venue standard error & Int effect est & Int se & Int var & Resid Var \\
      \midrule
      Older Finals & 0.0439 & $-1.901$ & 0.0251 & 0.0019 & 0.0109 \\
      All Finals & 0.0504 & $-1.914$ & 0.0280 & 0.0025 & 0.0112 \\
      Older Pooled & 0.0495 & $-1.853$ & 0.0305 & 0.0024 & 0.0243 \\
      All Pooled & 0.0554 & $-1.869$ & 0.0348 & 0.0031 & 0.0235 \\
      \bottomrule
  \end{tabular}
  \label{tab:Gamma_parameters}
\end{table}

At a size of $n=10,000,000$, the calculated probability of a reaction
time being below 0.1 seconds was as follows for the various different data sets:
\jy{label the table and reference to it; clean it up}
\begin{table}
  \centering
  \caption{Probabilities of observing reaction times less than 0.1, 0.09, and
  0.08 for various data sets.}
  \begin{tabular}{c c c c} 
   \toprule
   Data Set & P(RxnTime $<$ 0.1) & P(RxnTime $<$ 0.09) & P(RxnTime $<$ 0.08) \\ 
   \midrule
   all finals & $8.56\cdot10^{-4}$ & $4.1\cdot10^{-5}$ & $6.0\cdot10^{-7}$ \\ 
   old finals & $4.39\cdot10^{-4}$ & $1.63\cdot10^{-5}$ & $2.0\cdot10^{-5}$ \\
   all pooled & $6.86\cdot10^{-3}$ & $1.21\cdot10^{-3}$ & $1.3\cdot10^{-4}$ \\
   old pooled & $5.76\cdot10^{-3}$ & $9.93\cdot10^{-4}$ & $1.0\cdot10^{-4}$\\
   \bottomrule
  \end{tabular}
  \label{tab:Sim_probability}
\end{table}

@@

\jy{motivate first; describe the data (what are the grouping variables in the
  two data sets? use one paragraph for each.}
\of{Shouldnt describing the grouping variables go more in the data or methods 
section?  Currently, there is some writing about this but it is commented out. It
can be found in section 3.2}
